Centering factor interactions

### 2014-07-08 _knitr-posts/2014-06-12-centering-interactions.Rmd
Add draft version of interaction blog post

---
layout: post
title: Centering factor variables with interactions
subtitle: Why centering factor predictors can be important
published: false
---

@@

Notice how the main effects (everything except the interaction) change dramatically when the interaction is removed. This is because when the interaction is included the main effects are relevant to when the predictors are equal to 0. I.e. `x1 = 0` and the binary predictor `x2` is at its reference (0) level. But, when the interaction is excluded the main effects are relevant across all the values of the other predictors, or equivalently, their mean values. So, if we center, the main effects will represent the same thing in both cases:

@@

@@

```{r}
m2 <- lm(y ~ x1 * f, data = dat)
round(coef(m2), 2)
```

@@

```{r}
m2.1 <- lm(y ~ x1 * x2 + x1 * x3, data = dat)
round(coef(m2.1), 2)
```

@@

Now, let's compare to a model without the interaction:

```{r}
m2.1_no_inter <- lm(y ~ x1 + x2 + x1 + x3, data = dat)
round(coef(m2.1), 2)
round(coef(m2.1_no_inter), 2)
```

Again, if we model averaged here across models with and without the interaction, we'd be getting gibberish.

@@

```{r}
round(coef(m2.2), 2)
round(coef(m2.2_no_inter), 2)
```

@@

One more thing. Say we wanted to plot the predicted response from our model. This isn't quite as simple as normal once we've centered our predictors and coded the factor levels as centered dummy variable comparisons.

We'll make predictions at each of the 3 levels. Note how we create these levels through the 3 valid combinations of our dummy variables. We can't just use `expand.grid()` and get all combinations of our dummy variables.

@@

We can calculate the combination of these two distributions multiplied by the constants `a` and `b` (which we'll call `N3`) as:

@@

First, to illustrate combining the coefficient means, here's an example plotting the predictions by specifying the intercept and slopes. The first line in each of these `abline` function calls represents the intercepts and each factor level and the second line represents the slopes. Notice how we get these slopes by turning the two indicator columns to low or high values.

@@

>>>>

---
layout: post
title: Centering factor interactions
subtitle: Why and how you must center interactions for model averaging
published: false
---

@@

Notice how the main effects (everything except the interaction) change dramatically when the interaction is removed. This is because when the interaction is included the main effects are relevant to when the other predictors are equal to 0. I.e. `x1 = 0` or the binary predictor `x2` is at its reference `0` level. But, when the interaction is excluded the main effects are relevant when the other predictors are at their mean values. So, if we center, the main effects will represent the same thing in both cases:

@@

Here's an illustration:

<img src="/images/interactions.png" alt="Illustration of centering interactions" width=580 height=436>

@@

```{r}
m2 <- lm(y ~ x1 * f, data = dat)
t(round(coef(m2), 2))
```

@@

```{r}
m2.1 <- lm(y ~ x1 * x2 + x1 * x3, data = dat)
t(round(coef(m2.1), 2))
```

@@

Now, let's compare to a model without the interaction before we center the data:

```{r}
m2.1_no_inter <- lm(y ~ x1 + x2 + x1 + x3, data = dat)
t(round(coef(m2.1_no_inter), 2))
```

The main effects look dramatically different. Again, if we model averaged here across models with and without the interaction, we'd be getting gibberish.

@@

```{r}
t(round(coef(m2.2), 2)) # no interaction 
t(round(coef(m2.2_no_inter), 2)) # with interaction
```

@@

Say we wanted to plot the predicted response from our model on new data. This isn't quite as simple as normal once we've centered our predictors and coded the factor levels as centered dummy variable comparisons.

We'll make predictions at each of the 3 levels. Note how we create these levels through the 3 valid combinations of high and low values of our dummy variables. We can't just use `expand.grid()` and get all combinations of our dummy variables.

@@

We can calculate the combination of these two distributions multiplied by the constants `a` and `b` as: (we'll this new distribution `N3`)

@@

First, to illustrate combining the coefficient means, here's an example plotting the predictions by specifying the intercept and slopes. The first line in each of these `abline` function calls represents the intercepts and each factor level and the second line represents the slopes. As before, notice how we get these slopes by turning the two indicator columns to low or high values.

@@

Whether or not you center your predictors, transforming your factor coefficient interactions like the above can be a powerful way of displaying data. For example see Figure 1 in [O'Regan et al. 2014, Ecology, 95(4)](http://www.esajournals.org/doi/pdf/10.1890/13-0916.1) where we use this method to illustrate the effect of climate warming and drying on four responses across three amphibian species. Each panel represents a single model and we've adjusted the coefficients so that each effect is with respect to zero instead of with respect to another factor level. If this kind of presentation of the results is all you're after, then there's no need to center the factor levels. If, however, you also want to apply model averaging to coefficient values that include interactions, you will need to center your predictors and combine the coefficients as in the above example.
